Modelling BCG vaccination in the UK

### 2019-06-18 rmds/08-model-development.Rmd
updated model dev

# Devoloping a dynamic transmission model of Tuberculosis {#model-development}

@@

load_formated("disease_params_table")
load_formated("demographic_params_table")
load_formated("sources_table")
load_formated("eff_bcg")
```

@@

Non-UK born pulmonary cases were estimated using the ETS (see Chapter \@ref(data)) for each age-group included in the model from 2000 until 2015. Prior to 2000 a linear scale up of non-UK born cases, starting in 1960 and reaching the average number of cases observed in 2000 by December 1999 was assumed. A log link scale (i.e $log(t - t_{start})/log(t_{max} - t_{start})$) is also explored as a scenario during model fitting. To incorperate the uncertainty in the number of non-UK born cases a normal distribution was used, with the standard deviation determined using parameters from the observation model (Chapter \@ref(model-fitting)).

@@

\begin{equation}
\beta_a = \frac{(\nu_{\text{avg}}^{P} + \mu_{\text{avg}})\left(C_{\text{eff}} + C^{\text{hist}}_{\text{eff}} * t_{\text{hist}}\left(1 - \frac{log(t-1931)}{log(1990-1931)}\right)\right)}{C_{\text{actual}}}
(\#eq:beta-eq)
\end{equation}

@@

Vynnycky et al. found that the effective contact rate (i.e the transmission probability times the contact rate) for TB was approximately 22 in 1900 and fell to approximately 1 in 1990.[@Vynnycky1999] Incidence rates have increased since 1990 (Chapter \@ref(data)) and it is unclear what impact this has had on the effective contact rate. I have assumed that the effective contact rate is uniformly distributed between 0 and 5 contacts. Similarly, for the historic effective contact rate I have assumed a uniform distribution between 10 and 15. This gives a maximum effective contact rate of 20, inline with Vynnycky et al.'s estimate for the effective contact rate in 1900. I have also assumed that the historic contact rate declines over time using an exponential model, reducing to 0 in 1990. The prior for the half life of this decline ($c^{\text{hist}}_{\text{half}}$) has been assumed to be uniformly distributed between 0 and 20 years. Additional age stratification of $\beta$ is explored by including modifiers for certain age-groups. The baseline scenario is that no modification is required, with variation between children and adults ($\beta_{\text{child}}$) and between children, adults and older adults ($\beta_{\text{child}}$ and $\beta_{\text{older-adults}}$) also being considered. The priors for these modifiers are normally distributed around 1 with a standard deviation of 0.5 (truncated to be greater than 0). The contact rate is estimated by averaging the total age-specific contact rates estimated from POLYMOD data (Section \@ref(contact-matrix)) on an annual basis.

@@

The rate of recovery from active TB disease was estimated as 1 over the time with active disease using the ETS with UK born cases from 2000 until 2012. Cases with a period of time symptomatic that was less than 0 days were removed as these are likely to be spurious. Figure \@ref(fig:tb-time-treat) indicates that the distribution of time to treatment differs between children and adults and by pulmonary/extra-pulmonary TB status. There was little evidence that time to treatment differed between adults and older adults. A normal distribution was used for each age group, truncated to be greater than 0 months. Prior to 1952, and the introduction of isoniazid, I have assumed that the time to recovery from active TB disease is 20 years, representing natural recovery.

@@

The rate successful treatment was estimated as 1 over the period of time on treatment using the ETS, with UK born cases between 2000 and 2012. Cases with a treatment time less than 1 month were removed as TB treatment is standardised and should take at least several months. There was little evidence that time to treatment completion differed between pulmonary and extra-pulmonary TB cases but there was some evidence that older TB cases were more likely to be on treatment for longer than younger cases (Figure \@ref(fig:tb-treat-time-succ)). A normal distribution was used for children, adults and older adults, with each truncated to be greater than 4 months. This truncation was introduced as a faster treatment time than this was considered implausible.

@@

The demographic model paramters are outlined in Table \@ref(tab:demo-model-params), additional details are given in the following section. Table \@ref(tab:sources-tab) contains details of the sources used to parameterise the demographic model, again more detail is given in the following section for complex parameters.

@@

Model simulations were begun in 1931 due to the availability of population data from the 1931 census and because data on live births was only available until 1929. The demographic model is initialised using the age grouped 1931 census data, with the assumption that the entirety of the population is UK born. Initially it is assumed that there is no BCG vaccination and no TB treatment. This means that the coverage of BCG vaccination is zero. The rate of recovery from active disease is to be a minimum of 1/20, this reflects natural clearance of TB. TB treatment is assumed to begin in 1952 with the discovery of isoniazid and BCG vaccination begins at school-age in 1953. BCG vaccination coverage is assumed to vary randomly over the time horizon of the model but to have the same distribution. The assumed distribution is normal with a mean of 80% and a standard deviation of 5%. The duration with active TB is assumed to decrease from the introduction of treatment in 1952 through to 1990 when it is assumed that detection rates were equivalent to those seen today. I have assumed the following logistic model for the increase in the rate of starting treatment,

\begin{equation}
\nu_a(t) = \nu_a \frac{1}{1 + exp(\nu_{scale} * (t - 1952) - (1990 - 1952) / 2}
\frac{log(t - 1952)}{log(1990 - 1952)}
  (\#eq:adjusted-treatment-starting)
\end{equation}

This leaves $\nu_{scale}$ as a free scaling parameter.

@@

The model is initialised with the number of pulmonary and extra-pulmonary cases reported in 1931. To account for possible measurement error a normal distribution is sampled around the number of cases reported with a standard deviation of 5% of the reported cases. Cases are distributed across age groups based on the age distribution of the population.

@@

In the next chapter, I will consider an age modified transmission probablity, and the change in historic non-UK born cases over time as distinct scenarios (Table \@ref(tab:summary-scenarios)). Identifying if the transmission probablity varies with age is important as it may alter the distribution and number of TB cases. This would impact the observed effects of the BCG vaccine and is therefore of primary importance. The historic number of non-UK born cases is likely to be less translatable to other models, or scenarios, but is important to fully investigate here as it determines the proportion of the population with different stages of latent TB in 2000 when data on TB notifications is first available. This is important as a higher proportion of latent TB cases in the population leads to a greater number of re-activations. This in turn results in a reduction in TB transmission when fitting to observed incidence.

```{r summary-scenarios}
scenarios <- tibble(
  Parameter = c("$\\beta_a$ - transmission probability", "", ""),
  Scenario = c("Constant across age groups",
               "Variable between children (0-14) and adults (15-89)",
               "Variable between children (0-14), adults (15-59), and older adults (60-80)")
) %>% 
  add_row(
    Parameter = c("Historic non-UK born cases", "", ""),
    Scenario = c("Increasing from 1960 to 1999 based on notifications in 2000 using a log link function.",
                 "Linear increase from 1960 to 1999 based on notifications in 2000",
                 ""
                 )
  )

kable(scenarios, caption = "Summary of planned scenario analyses to be carried out in the next chapter as part of model fitting by comparision of the goodness of fit to the data.", booktabs = TRUE, escape = FALSE) %>% 
  column_spec(c(1), width = "2cm") %>% 
  column_spec(c(2), width = "8cm")
```

This chapter has outlined a realistic dynamic transmission model of TB that includes key features required to investigate BCG vaccination policy and is robustly parameterised from an extensive, and previously unused in a TB model, routine surveillance data-set. Transformations, and approximations, of parameters have been used to make the best use of available data. However, there are several key limitations. Firstly, the model presented here does not explicitly model TB transmission in the non-UK born. This means that in order to initialise the model assumptions must be made about the historic number of non-UK born cases and the future incidence in the non-UK born must also be assumed in order to produce projections of future TB incidence. However, this simplification allows many complexities of TB in the non-UK born to be discounted, such as the rate of case importation, heterogeneity amongst the non-UK born from different countries, and mixing within the non-UK born. It also means that, as stated above, explicit assumptions must be made about the number of historic, and future, TB cases in the non-UK born. This makes the impact of these assumptions clearer as a scenario analysis can be conducted, whereas if non-UK born cases were explicitly modeled incidence in these populations would depend on more abstract assumptions. Finally, the majority of the non-UK born population would not be exposed to England's BCG vaccination policy and so are unlikely to be impacted by changes in this policy. Secondly, the model presented here does not include high and low risk stratification within the UK born. Individuals that are from countries with incidence above 40 per 100,000, or that have parents/grandparents from countries with incidence above 40 per 100,000, are considered at higher risk of TB.[@PublicHealthEngland2011a] In addition, individuals living in areas of the UK with incidence above this threshold are also considered at higher risk. Current BCG vaccination policy targets high risk neonates for vaccination, with low risk neonates not being vaccinated. Ideally this high/low risk stratification would be included as it would allow the evaluation of the current BCG vaccination policy. Unfortunately this has not been possible as there is little data from which to extrapolate either the number of high risk notifications in the ETS, or the size of the high risk population. There is also little evidence to suggest the degree of mixing between the non-UK born, the high risk UK born population and the low risk non-UK born population. It is likely that introducing this structure into the model, without the data outlined above, would lead to the model being poorly specified and therefore failing to fit to the observed data. Instead, in the final chapter in this thesis, the high risk neonatal programme will be proxied by a universal neonatal programme. This will allow for a comparison to be made between school-age and neonatal vaccination but does not allow for the impact of targeting high risk individuals to be evaluated. Finally, the model presented here does not include the full complexity of TB epidemiology. Drug resistant TB may have more severe outcomes, standard TB treatment may fail resulting in a longer period on treatment, and TB outcomes may vary by risk group.[@PHE2017] However, drug resistant TB cases are known to make up a small fraction of TB cases in England in the UK born, variable treatment times have been included in the prior distribution of treatment times and TB outcomes. Model parameters have also been stratified by pulmonary status and age group where appropriate. Additionally, complexity has been included for the action of the BCG vaccine, with realistic waning in effectiveness. Observed age-specific mortality rates and the number of live births has also been included, allowing for realistic population demographics. This means that estimates of the impact of the change in vaccination policy are likely to be more accurate, whereas a more complex model of the epidemiology of TB would likely have little impact on these results.

@@

Several studies have made use of dynamic TB transmission models to evaluate BCG vaccination or future vaccines.[@Harris2016; @Egbetade2011a; @Bhunu2008b] In general these studies used less detailed models than the one presented here, typically because they were modelling TB in a more generic setting or because more information about TB epidemiology, TB natural history, and the BCG vaccine has become available over time. In addition, no dynamic model of TB, including BCG vaccination, has currently been published that includes both protection from initial infection and protection from active TB due to BCG vaccination. There have also been no studies that use the current best estimates for the duration of BCG protection in developed countries away from the equator. Harris et al. reviewed mathematical models that explored the epidemiological impacts of future TB vaccines. They found that vaccines targeted at all-ages or at adolescents/adults were more effective at eradicating TB than neonatal programmes when vaccine effectiveness was not assumed to degrade with age. The majority of studies included in their review used deterministic, compartmental, dynamic models. Model structures were found to have evolved over time as TB natural history and epidemiology is better understood, with the majority of models having at least susceptible, latent, active disease, and recovered states. Treatment status, variable infectiousness of active disease, vaccine waning, and age stratification were included in some of the models evaluated.[@Harris2016] Recently it has been shown that only models that include at least two latent compartments are able to reproduce the observed activation dynamics of TB.[@Ragonnet2017] The model presented here is based on the serial latency archetype identified in this study. It has also been shown that realistic age structure and population demographics, included in the model presented in this chapter, are critical for reproducing TB epidemiology.[@Brooks-Pollock2010] Egbetade et al. presented a dynamic model of TB that included BCG vaccination but did not include age structure. They found that universal vaccination increased the stability of the disease free equilibrium in countries with high TB burden. However the model presented was not rigorously parameterised with data and only a single latent TB compartment was used.[@Egbetade2011a] Bhunu et al. developed a dynamic transmission model of TB that in order to investigate the effects of pre- and post-exposure vaccines for tuberculosis control. Again their model did not include multiple latent compartments or age structure unlike the model presented here.[@Bhunu2008b]

>>>>

# Developing a dynamic transmission model of Tuberculosis {#model-development}

@@

load_formated("disease_params_table")
load_formated("demographic_params_table")
load_formated("sources_table")
#load_formated("scenarios_table")
load_formated("eff_bcg")
```

@@

Non-UK born pulmonary cases were estimated using the ETS (see Chapter \@ref(data)) for each age-group included in the model from 2000 until 2015. Prior to 2000, incidence in the non-UK born are unavailable and the relationship to transmission in UK-born cases is unknown. To account for this, importation of non-UK born cases in the model begins in 1960 and then is scaled up through to 2000. As the form of this relationship is unknown the following functional form was used to scale cases based on those observed in 2000,

\begin{equation}
\text{Non UK-born scaling} = \frac{e^{\left(\frac{t - 1960}{\text{ln}(2) * \iota_{\text{scale}}}\right)} - 1}{e^{\left(\frac{2000 - 1960}{\text{ln}(2) * \iota_{\text{scale}}}\right)} - 1}
  (\#eq:non-uk-born-scale)
\end{equation}

This functional form was chosen as it is flexible enough to represent exponential growth, log constrained growth and near linear growth depending on the choice of $\iota_{\text{scale}}$. This allows the scale up of non-UK born cases to be fitted to the available data during the model fitting stage (Chapter \@ref(model-fitting)). To incorperate the uncertainty in the number of observed non-UK born cases a normal distribution was used, with the standard deviation and mean determined using parameters from the observation model (Chapter \@ref(model-fitting)).

@@

\begin{equation}
C^{\text{scaled}}_{\text{eff}} = \begin{cases}
C^{\text{hist}}_{\text{eff}} & \text{if } t < 1950, \\
C_{\text{eff}} + \left(C^{\text{hist}}_{\text{eff}} -  C_{\text{eff}} \right) \left(1 - \frac{t-1950}{C^{\text{year}}_{\text{eff}}-1950}\right) & \text{if } 1950 \leq t \leq 1990, \\
C_{\text{eff}} & \text{if } t > 1990. \end{cases}
  (\#eq:scaled-contacts)
\end{equation}

\begin{equation}
\beta_a = \frac{(\nu_{\text{avg}}^{P} + \mu_{\text{avg}}) C^{\text{scaled}}_{\text{eff}}}{C_{\text{actual}}}
(\#eq:beta-eq)
\end{equation}

@@

Vynnycky et al. found that the effective contact rate (i.e the transmission probability times the contact rate) for TB was approximately 22 in 1900 and fell to approximately 1 in 1990.[@Vynnycky1999] Incidence rates have increased since 1990 (Chapter \@ref(data)) and it is unclear what impact this has had on the effective contact rate. I have assumed that the effective contact rate is normally distributed with a mean of 1 and a standard deviation of 0.5. Similarly, for the historic effective contact rate I have assumed a normal distribution between with a mean of 10 and a standard deviation of 2.5. I have also assumed that the historic contact rate declines linearly over time, starting in 1950, reducing to 0 by $C^{\text{year}}_{\text{eff}}$. The prior for this year has been assumed to be normally distributed with a mean of 1985, and a standard deviation of 5 years. It has been truncated so that the lower bound is 1975 and the upper bound is 1995. Additional age stratification of $\beta$ is explored by including modifiers for certain age-groups. The baseline scenario is that no modification is required, with variation between children and adults ($\beta_{\text{child}}$) and between children, adults and older adults ($\beta_{\text{child}}$ and $\beta_{\text{older-adults}}$) also being considered. The priors for these modifiers are normally distributed around 1 with a standard deviation of 0.5 (truncated to be greater than 0). The contact rate is estimated by averaging the total age-specific contact rates estimated from POLYMOD data (Section \@ref(contact-matrix)) on an annual basis.

@@

The rate of recovery from active TB disease was estimated as the reciprocal of the time with active, untreated, disease from the ETS with UK born cases from 2000 until 2012. Cases with a period of time symptomatic that was less than 0 days were removed as these are likely to be spurious. Figure \@ref(fig:tb-time-treat) indicates that the distribution of time to treatment differs between children and adults and by pulmonary/extra-pulmonary TB status. There was little evidence that time to treatment differed between adults and older adults. A normal distribution was used for each age group, truncated to be greater than 0 months. Prior to 1952, and the introduction of isoniazid, I have assumed that the time to recovery from active TB disease is 2 years, representing natural recovery or other removal from the infectious population. From 1952 to 1990 the time with active TB is assumed to decrease linearly.

@@

The rate successful treatment was estimated as the reciprocal of the period of time on treatment using the ETS with UK born cases between 2000 and 2012. Cases with a treatment time less than 1 month were removed as TB treatment is standardised and should take at least several months. There was little evidence that time to treatment completion differed between pulmonary and extra-pulmonary TB cases but there was some evidence that older TB cases were more likely to be on treatment for longer than younger cases (Figure \@ref(fig:tb-treat-time-succ)). A normal distribution was used for children, adults and older adults, with each truncated to be greater than 4 months. This truncation was introduced as a faster treatment time than this was considered implausible.

@@

The demographic model parameters are outlined in Table \@ref(tab:demo-model-params), additional details are given in the following section. Table \@ref(tab:sources-tab) contains details of the sources used to parameterise the demographic model, again more detail is given in the following section for complex parameters.

@@

Model simulations were begun in 1931 due to the availability of population data from the 1931 census and because data on live births was only available until 1929. The demographic model is initialised using the age grouped 1931 census data, with the assumption that the entirety of the population is UK born. Initially it is assumed that there is no BCG vaccination and no TB treatment. TB treatment is assumed to begin in 1952 with the discovery of isoniazid and BCG vaccination begins at school-age in 1953. BCG vaccination coverage is assumed to vary randomly over the time horizon of the model but to have the same distribution at all time points. The assumed distribution is normal with a mean of 80% and a standard deviation of 5%. The duration with active TB is assumed to decrease from the introduction of treatment in 1952 through to 1990 when it is assumed that detection rates were equivalent to those seen today.

@@

The model is initialised with the number of pulmonary and extra-pulmonary cases reported in 1931. The high risk latent population is initialised by scaling the number of observed cases in 1931 by the proportion of high risk latent cases that develop active TB, the duration that these cases are high risk and then dividing by the infectious period. The low risk latent population is then initialised by scaling the high risk latent population linearly by average age in compartment. Finally, the initial susceptible population is based on the population estimate from the 1931 census minus the assumed initial latent cases.

All initial disease compartments are distributed based on the age distribution of observed UK born cases in 2000. To account for possible measurement error a normal distribution is sampled around the assumed population in each compartment with a standard deviation of 5% of the reported cases.

@@

In the next chapter, I will consider the evidence for modifying the transmission probability by age using three distinct scenarios (Table \@ref(tab:summary-scenarios)). Identifying if the transmission probablity varies with age is important as it may alter the distribution and number of TB cases. This would impact the observed effects of the BCG vaccine and is therefore of primary importance.

```{r summary-scenarios, eval = FALSE}
kable(scenarios_table, caption = "Summary of planned scenario analyses to be carried out in the next chapter as part of model fitting by comparision of the goodness of fit to the data.", booktabs = TRUE, escape = FALSE) %>% 
  column_spec(c(1), width = "2cm") %>% 
  column_spec(c(2), width = "8cm")
```

This chapter has outlined a realistic dynamic transmission model of TB that includes key features required to investigate BCG vaccination policy and is robustly parameterised from an extensive, and previously unused in a TB model, routine surveillance data-set. Transformations, and approximations, of parameters have been used to make the best use of available data. However, there are several key limitations. Firstly, the model presented here does not explicitly model TB transmission in the non-UK born. This means that in order to initialise the model assumptions must be made about the historic number of non-UK born cases and the future incidence in the non-UK born must also be assumed in order to produce projections of future TB incidence. However, this simplification allows many complexities of TB in the non-UK born to be discounted, such as the rate of case importation, heterogeneity amongst the non-UK born from different countries, and mixing within the non-UK born. Secondly, the model presented here does not include high and low risk stratification within the UK born. Individuals that are from countries with incidence above 40 per 100,000, or that have parents/grandparents from countries with incidence above 40 per 100,000, are considered at higher risk of TB.[@PublicHealthEngland2011a] In addition, individuals living in areas of the UK with incidence above this threshold are also considered at higher risk. Current BCG vaccination policy targets high risk neonates for vaccination, with low risk neonates not being vaccinated. Ideally, this high/low risk stratification would be included as it would allow the evaluation of the current BCG vaccination policy. This has not been possible as there is little data from which to extrapolate either the number of high risk notifications in the ETS, or the size of the high risk population. There is also little evidence to suggest the degree of mixing between the non-UK born, the high risk UK born population and the low risk non-UK born population. It is likely that introducing this structure into the model, without the data outlined above, would lead to the model being poorly specified and therefore failing to fit to the observed data. Instead, in the final chapter in this thesis, the high risk neonatal programme will be proxied by a universal neonatal programme. This will allow for a comparison to be made between school-age and neonatal vaccination but does not allow for the impact of targeting high risk individuals to be evaluated. Finally, the model presented here does not include the full complexity of TB epidemiology. Drug resistant TB may have more severe outcomes, standard TB treatment may fail resulting in a longer period on treatment, and TB outcomes may vary by risk group.[@PHE2017] However, drug resistant TB cases are known to make up a small fraction of TB cases in England in the UK born, and variable treatment times have been included in the prior distribution of treatment times and TB outcomes. Model parameters have also been stratified by pulmonary status and age group where appropriate. Additionally, complexity has been included for the action of the BCG vaccine, with realistic waning in effectiveness. Observed age-specific mortality rates and the number of live births has also been included, allowing for realistic population demographics. This means that estimates of the impact of the change in vaccination policy are likely to be more accurate, whereas a more complex model of the epidemiology of TB would likely have little impact on these results.

@@

Several studies have made use of dynamic TB transmission models to evaluate BCG vaccination or future vaccines.[@Harris2016; @Egbetade2011a; @Bhunu2008b] In general, these studies used less detailed models than the one presented here, typically because they were modelling TB in a more generic setting or because more information about TB epidemiology, TB natural history, and the BCG vaccine has become available over time. In addition, no dynamic model of TB, including BCG vaccination, has currently been published that includes both protection from initial infection and protection from active TB due to BCG vaccination. There have also been no studies that use the current best estimates for the duration of BCG protection in developed countries away from the equator. Harris et al. reviewed mathematical models that explored the epidemiological impacts of future TB vaccines. They found that vaccines targeted at all-ages or at adolescents/adults were more effective at eradicating TB than neonatal programmes when vaccine effectiveness was not assumed to degrade with age. The majority of studies included in their review used deterministic, compartmental, dynamic models. Model structures were found to have evolved over time as TB natural history and epidemiology is better understood, with the majority of models having at least susceptible, latent, active disease, and recovered states. Treatment status, variable infectiousness of active disease, vaccine waning, and age stratification were included in some of the models evaluated.[@Harris2016] Recently it has been shown that only models that include at least two latent compartments are able to reproduce the observed activation dynamics of TB.[@Ragonnet2017] The model presented here is based on the serial latency archetype identified in this study. It has also been shown that realistic age structure and population demographics, included in the model presented in this chapter, are critical for reproducing TB epidemiology.[@Brooks-Pollock2010] Egbetade et al. presented a dynamic model of TB that included BCG vaccination but did not include age structure. They found that universal vaccination increased the stability of the disease free equilibrium in countries with high TB burden. However the model presented was not rigorously parameterised with data and only a single latent TB compartment was used.[@Egbetade2011a] Bhunu et al. developed a dynamic transmission model of TB that in order to investigate the effects of pre- and post-exposure vaccines for tuberculosis control. Again their model did not include multiple latent compartments or age structure unlike the model presented here.[@Bhunu2008b]
